set---
title: "Analysis"
author: "Emma Borsier"
date: "23/02/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE)
```

```{r import_libraries, echo=FALSE}
library("readxl")
library("ggplot2")
```

## Load the dataset

```{r import_data}
df_csv <- read.csv("../data/raw_data.csv")
```

We remove the rows with missing values in the 'Entry' column, then convert the categorical 'Treatment'  column to a factor with four levels (for each of the four brightness conditions).
```{r preprocessing}
df <- subset(df_csv,!(is.na(Entry)))
df$Treatment <- factor(df$Treatment, levels=unique(df$Treatment))

```


We then make subsets of the complete dataset for each of the four brightness conditions / Treatments:
```{r subsets}
right_bright_df <- subset(df, Treatment=="Dark-Bright")
left_bright_df <- subset(df, Treatment=="Bright-Dark")
bright_df <- subset(df, Treatment=="Bright-Bright")
dark_df <- subset(df, Treatment=="Dark-Dark")

```

## Histograms

We plot the distribution of the horizontal position of the birds as they first enter the tunnel for the whole dataset, then for each of the four brightness conditions individually. \
On the x-axis, -1 represents the left edge, 0 the middle, and 1 the right edge of the tunnel entrance. 
```{r histograms, echo=FALSE}

hist(df$Entry,
     main="All conditions", 
     xlab="Horizontal position",
     breaks=10,
     xlim=c(-1,1),
     ylim=c(0,100),
     las=1)
hist(bright_df$Entry,
     main="Bright (symmetric)", 
     xlab="",
     breaks=10,
     col="#FFFC90",
     xlim=c(-1,1),
     ylim=c(0,25),
     las=1)
hist(dark_df$Entry,
     main="Dark (symmetric)", 
     xlab="Entry",
     breaks=6,
     col="#585858",
     xlim=c(-1,1),
     ylim=c(0,25),
     las=1)
hist(right_bright_df$Entry,
     main="Bright on the right", 
     xlab="",
     col="#97D4C4",
     breaks=10,
     xlim=c(-1,1),
     ylim=c(0,25),
     las=1)
hist(left_bright_df$Entry,
     main="Bright on the left", 
     xlab="",
     breaks=10,
     col="#EF85E5",
     xlim=c(-1,1),
     ylim=c(0,25),
     las=1)

```


## Modality

The distributions shown above do not appear to all be be unimodal, thus, we conducted a Hartigan's dip test to test for unimodality of the data for each brightness condition. 
```{r modality}

library(diptest) # package used to compute Hartigan's dip test 
library(multimode) # package used to compute the mode(s) of the entry point, which is a continuous variable 

dip.test(bright_df$Entry)
bright_modes <- locmodes(bright_df$Entry, mod0 = 2, display=TRUE, addplot=FALSE, addLegend=FALSE)
bright_modes

dip.test(dark_df$Entry)
dark_modes <- locmodes(dark_df$Entry, mod0 = 2, display=TRUE, addplot=FALSE, addLegend=FALSE)
dark_modes

dip.test(left_bright_df$Entry)
right_bright_modes <- locmodes(right_bright_df$Entry, mod0 = 2, display=TRUE, addplot=FALSE, addLegend=FALSE)
right_bright_modes

dip.test(right_bright_df$Entry)
left_bright_modes <- locmodes(left_bright_df$Entry, mod0 = 1, display=TRUE, addplot=FALSE, addLegend=FALSE)
left_bright_modes

```


## Effects of brightness condition on Origin side

```{r Origin table}
# We exclude the ambiguous data points for Origin from the analysis
Origin_table <- with(subset(df, Origin!='?'), table(Treatment, Origin))
Origin_table

```

We use this table to compute the proportions of Left and Right origin for each brightness Treatment. We then find the confidence intervals for these proportions using the method of Agresti and Coull, computed in Excel. We import the lower and upper bounds of these intervals from the Excel file named 'barplots_CI_tables.xlsx':

```{r import_ori_table}
origin_freq_df <- read_excel("barplots_CI_tables.xlsx", sheet = 2)
origin_freq_df$Treatment <- factor(origin_freq_df$Treatment, 
                                   levels=c("Left", "Right", "Bright", "Dark"))
```

We plot the proportions of flights originating from the left vs right (relative to the total of recorded flights), with the confidence intervals obtained above. 

```{r plot_ori_freq, echo=FALSE}
ggplot(origin_freq_df, aes(x=Treatment, y=Frequency, fill = Origin)) +
      geom_bar(stat="identity", position='dodge')+
      coord_cartesian( ylim = c(0, 100))+
      labs(x = "Brightness condition", y = 'Relative frequencies (%)') +
      scale_fill_manual(values = c("#F89933", "#F5D0A9"))+
      geom_errorbar( aes(ymin=Freq_lower, ymax=Freq_upper),
                     position=position_dodge(0.9), width=.15,)
```


## Effects of brightness condition on Arrival side

```{r arrival table}
# We exclude the ambiguous data points for Arrival from the analysis
Arrival_table <- with(subset(df, Arrival!='?'), table(Treatment, Arrival))
Arrival_table
```

Similarly, we use the above table to compute the proportions of Left and Right arrivals for each brightness Treatment. Once again, we then find the confidence intervals for these proportions using the method of Agresti and Coull, computed in Excel. We import the lower and upper bounds of these intervals from the same Excel file:

```{r import_arr_table}
arrival_freq_df <- read_excel("barplots_CI_tables.xlsx", sheet = 1)
arrival_freq_df$Treatment <- factor(arrival_freq_df$Treatment, 
                                   levels=c("Left", "Right", "Bright", "Dark"))
```

We plot the proportions of flights arriving on the left vs right (relative to the total of recorded flights), with the confidence intervals obtained above. 

```{r plot_arr_freq, echo=FALSE}
ggplot(arrival_freq_df, aes(x=Treatment, y=Frequency, fill = Arrival)) +
      geom_bar(stat="identity", position='dodge')+
      coord_cartesian( ylim = c(0, 100))+
      labs(x = "Brightness condition", y = 'Relative frequencies (%)') +
      scale_fill_manual(values = c("#2995d6","#9FC9EB")) +
      geom_errorbar( aes(ymin=Freq_lower, ymax=Freq_upper),
                     position=position_dodge(0.9), width=.15,)
```

## Effect of brightness condition on entry after grouping data by origin

```{r brightness boxplot, echo=FALSE}

q <- ggplot(subset(df, (Origin!='?' & Arrival!='?')), 
            aes(x=Origin, y=Entry, fill=Treatment)) + 
  geom_boxplot(alpha = 1) + scale_fill_manual(values = c("#EF85E5", "#97D4C4","#FFFC90","#716F6F")) +
  scale_y_continuous(limits = c(-1,1)) 

contrast_boxplot <- q + theme(
  panel.background = element_rect(fill='transparent'),
  plot.background = element_rect(fill='transparent', color=NA),
  panel.grid.major = element_blank(), #element_line(color='grey'),
  panel.grid.minor = element_blank(),
  legend.background = element_rect(fill='transparent'),
  legend.box.background = element_rect(fill='transparent'),
) 
contrast_boxplot
```


## Effect of arrival side on entry after grouping data by origin


```{r arrival boxplot, echo=FALSE}
q <- ggplot(subset(df, (Origin!='?' & Arrival!='?')), aes(x=Origin, y=Entry, fill=Arrival)) + 
  geom_boxplot(alpha=1)+ scale_fill_manual(values = c("#2995d6","#9FC9EB"))+
  scale_y_continuous(limits = c(-1,1)) 

ori_boxplot <- q + theme(
  panel.background = element_rect(fill='transparent'),
  plot.background = element_rect(fill='transparent', color=NA),
  panel.grid.major = element_blank(), #element_line(color='grey'),
  panel.grid.minor = element_blank(),
  legend.background = element_rect(fill='transparent'),
  legend.box.background = element_rect(fill='transparent'),
) 
ori_boxplot

```

# Statistical analysis

## Check for normality of data distribution


```{r plot_fn, echo=FALSE}

plot_residuals <- function(model) {
  res <- resid(origin_lm)
  plot(fitted(origin_lm), res)
  qqnorm(res)
  qqline(res)
  plot(density(res))
}
```

First, we look at the distribution of residuals for the simplest linear model: Entry ~ Origin

```{r normality_lm1}
origin_lm <- lm(Entry ~ Origin, data = df)
plot_residuals(origin_lm)
```

Then, we look at the distribution of residuals for the following two way ANOVA (in Statistical Analysis section):
Entry ~ Origin*Treatment

```{r normality_lm2}
ori_treatment_lm <- lm(Entry ~ Origin*Treatment, data = df)
plot_residuals(ori_treatment_lm)
```

Finally, we look at the distribution of residuals for the following two way ANOVA (in Statistical Analysis section):
Entry ~ Origin*Arrival

```{r normality_lm3}
arr_treatment_lm <- lm(Entry ~ Origin*Arrival, data = df)
plot_residuals(plot_residuals(ori_treatment_lm))
```

#### Conclusion:
The residuals of these 3 linear models are approximately normally distributed, but with a small number of outliers attributable to cases where the bird entered the tunnel on the opposite side to the model’s prediction


## Categorical data analysis

### Chi-squared test
Testing for an effect of the brightness condition on the preferred side of Origin
```{r chi_test}
# We exclude the ambiguous data points for Arrival from the analysis
Origin_table <- with(subset(df, Origin!='?'), table(Treatment, Origin))
Origin_table
chisq.test(Origin_table)

```

### Chi-squared test 
Testing for an effect of the brightness condition on the preferred side of Arrival
```{r chi_test2}
# We exclude the ambiguous data points for Arrival from the analysis
Arrival_table <- with(subset(df, Arrival!='?'), table(Treatment, Arrival))
Arrival_table
chisq.test(Arrival_table)
```

## Continuous data analysis

### Two-sample t-test 
Checking for any significant effect of Origin on the point of Entry
```{r t_test}
t.test(Entry ~ Origin, data = subset(df,Origin!='?'), var.equal = TRUE, paired = FALSE, alternative="two.sided")
```

```{r means_sd, echo=FALSE}

right_ori_df = subset(df,Origin=='Right')
n_right = nrow(right_ori_df)
mean_right = mean(right_ori_df$Entry)
sd_right = sd(right_ori_df$Entry)
sprintf("Birds originating from the right tended to enter to the right of centre (φ = %f ± %f; mean ± s.d; n = %i)", mean_right, sd_right, n_right)

left_ori_df = subset(df,Origin=='Left')
mean_left = mean(left_ori_df$Entry)
n_left = nrow(left_ori_df)
sd_left = sd(left_ori_df$Entry)
sprintf("Birds originating from the left tended to enter to the left of centre (φ = %f ± %f; mean ± s.d; n = %i)", mean_left, sd_left, n_left)
```


### Two-way ANOVAs

1. Checking for any significant interaction between the effect of Origin side and the effect of the brightness condition:

```{r anova1}
# 2-way ANOVA with Treatment (brightness) and Origin as factors + their interaction
anova_brightness <- aov(Entry ~ Origin*Treatment, data = subset(df,Origin!='?' & Arrival!='?'& (Treatment=='Dark-Bright' | Treatment=='Bright-Dark')))
summary(anova_brightness)
```

2. Checking for any significant interaction between the effect of Origin side and the effect of Arrival side:

```{r anova2}
# 2-way ANOVA with Arrival and Origin as factors + their interaction
anova_arrival <- aov(Entry ~ Origin*Arrival, data = subset(df, (Origin!='?' & Arrival!='?')))
summary(anova_arrival)

```



